rapr: Representative and Adequate Prioritisations in R

### 2016-06-23 code/rmarkdown/article.Rmd
+ major article edits to results and discussion + add in cs1 figs to si + fix bug in make clean

---
title: "rapr: Representative and Adequate Prioritisations in R"
author: |
  | Jeffrey O. Hanson$^1$, Jonathan R. Rhodes$^2$, Hugh P. Possingham$^1$, Richard A. Fuller$^1$
  | $^1$School of Biological Sciences, The University of Queensland, Brisbane, QLD, Australia
  | $^2$School of Geography, Planning and Environmental Management, The University of Queensland, Brisbane, QLD, Australia
  | Correspondance should be addressed to jeffrey.hanson@uqconnect.edu.au
date: "`r format(Sys.time(), '%d %B %Y')`"
tags: ["conservation", "biodiversity", "MILP", "reserve-selection"]
abstract: "A central aim in conservation is to maximise the long-term persistence of biodiversity. To fulfil this aim, reserve networks are used to safeguard biodiversity patterns (eg. species, populations) and processes (eg. evolutionary processes that underpin genetic variation). Reserve selection is often formulated as an optimisation problem to identify cost-effective prioritisations. However, most existing decision support tools are based on formulations that are well suited for preserving biodiversity patterns, but not biodiversity processes. To fill this gap in the conservation planning toolbox, we developed the \texttt{rapr R} package. This \texttt{R} package provides a toolkit to guide reserve selection using novel formulations of this problem. Here, we explore the functionality of this R package using a simulation study and two case-studies. We demonstrate how explicitly considering biodiversity processes can alter a prioritisation. In most cases, we found that only a few additional plannning units are required to sufficiently preserve of biodiversity processes."
output:
  rmarkdown::pdf_document:
    toc: false
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: preamble-latex.tex
fontsize: 11pt
documentclass: article
bibliography: references.bib
csl: reference-style.csl
vignette: >
  %\VignetteIndexEntry{rapr}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

@@

## Introduction
<!-- conservation and reserve networks -->
The overarching aim of conservation is to maximise the long-term persistence of biodiversity [@r423; @r255]. To achieve this, conservation actions must preserve biodiversity patterns (eg. species, populations) and the processes that sustain them. One of the major tangible achievements of modern conservation has been the act of setting aside areas for preservation [@r440]. Reserve networks buffer species from threatening processes [eg. urbanisation; @r255] and set the stage for direct management interventions [eg. captive breeding and reintroduction programs; @r290]. However, the resources available for conservation action are limited, and so reserve networks must be sited in places that satisfy conservation objectives for minimum cost [@r255]. To achieve this, reserve selection is often formulated as an optimisation problem and then solved to identify cost-effective candidate reserve systems [prioritisations; @r255].

<!-- biodiversity processes are important -->
To fulfil the overarching aims of conservation, reserve networks must preserve both ecological and evolutionary processes [@r3; @r255]. Ecological processes, such as predator-prey interactions, pollination, and decomposition, are required for biodiversity to persist over short time-scales. Typically, they operate over small geographic domains--with exceptions such as migration and refugial habitats--and can be preserved using suitably large planning units [@r446] that each contain a discrete unit of habitat [@r448]. On the other hand, evolutionary processes are required for biodiversity to persist over long time-scales, and they typically operate over large geographic domains. Protected areas must preserve adaptive evolutionary processes to foster resilience against environmental change [eg. climate change; @r247]. Protected areas must also preserve neutral evolutionary processes [@r5; @r67; @r63]. These processes are associated with the breakdown of gene flow between populations. They are important for maintaining genetic diversity, and avoiding inbreeding depression. In recent decades, a wealth of data on biodiversity processes has become available to conservation planners [eg. bioclimatic, genetic, and trait data; @r33; @r453; @r460]. Yet this data is only rarely used to guide conservation planning [@r61]. This is due to that fact that existing reserve selection tools focus on preserving biodiversity patterns or processes--but not both.

@@

<!-- aims of this paper -->
Today, one of the key issues in reserve selection is the lack of a unifying decision support tool that can accommodate data on biodiversity patterns and processes in a multi-species context. To fill this void, we present the \texttt{rapr R} package. This \texttt{R} package uses novel formulations of the reserve selection problem to provide decision makers with the tools to generate prioritisations that preserve biodiversity patterns and processes. We also explore the functionality of this \texttt{R} package using simulated species and a case-study conservation planning scenario.

## Problem formulations
The \texttt{rapr R} package uses two novel formulations of the reserve selection problem to identify cost-effective prioritisations. These formulations share many constraints and variables. For brevity, the variables used by both formulations will be defined. Biodiversity features are defined as the entity(s) that the prioritisation is required to preserve (eg. species, populations). Spatial attributes are defined as the variation across the species' range that the prioritisation is required to sample. They can be intrinsic (eg. genetic or trait variation) or extrinsic (eg. environmental variation) to the feature. These attributes should be related to the biodiversity processes that the decision maker aims to preserve.

@@

To illustrate these concepts, we will briefly describe a conservation planning scenario example involving an attribute spaces and demand points. A decision maker may wish to develop a prioritisation for a single species. This species has four populations in the study area. However, the decision maker can only afford to preserve three population. The decision maker needs to preserve the adaptive landscape of the species, and preserve populations inhabiting different environmental conditions. To describe environmental variaiton, the decision maker obtained data on the environmental conditions (rainfall (ml) and temperature ($^{\circ}$C)) where each population was found. The decision maker then used this environmental data to construct a two-dimensional environmental attribute space. Next, the decision maker generated demand points as equi-distant points between the range of values where the populations were found [$\pm$ 20% to avoid edge effects; @r218]. By comparing the distribution of the demand points to the distribution of the populations in the attribute space, the decision maker can identify a suitable prioritisation (Figure 1). We can see that if the decision maker preserves both populations $A$ and $C$, they will effectively "double-up" on the same environmental characteristics, and in turn their waste resources. Instead, a more representative sample of the intra-specific variation could be preserved by securing populations $A$, $B$, and $D$. This example demonstrates how the inclusion of biodiversity processes can guide the reserve selection process.

The problem formulations used to guide reserve selection in the \texttt{rapr R} package are based on a combination of the \texttt{Marxan} reserve selection problem and the uncapacitated facility location problems [@r451]. All mathematical terms defined hereafter are described in Table S1 for convenience. For convenience, the cardinality of sets will be denoted using the same symbol used to denote the variable. Define $F$ to be the set of features (indexed by $f$). Let $J$ be a set of planning units (indexed by $j$). Also, let $A_j$ denote the area, and $C_j$ denote the cost of preserving planning unit $j \in J$. To assess the extent to which each feature is secured in a given prioritisation, let $q_{fj}$ denote the probability of feature $f$ occupying planning unit $j$. The level of fragmentation associated with a prioritisation is parametrised as the net exposed boundary length. Let the shared edges between each planning unit $j \in J$ and $k \in J$ be $e_{jk}$.

@@

The \texttt{R} package provides two formulations for reserve selection. They are based on the unreliable [@r451] and reliable uncapacitated facility location [@r16] problems. The key difference between them is that the reliable formulation explicitly considers the probability that the planning units are occupied when calculating the proportion of variation that a given solution secures, whereas the unreliable formulation does not.

### Unreliable formulation
In the unreliable formulation, the control variables are the $B$, $T_{s}$, and $\tau_{sa}$ variables.

@@

The objective function (3a) determines the utility of a given prioritisation: a combination of the total cost of a prioritisation and how fragmented it is. Constraints (3b) ensure that all the amount-based targets are met. Constraints (3c) ensure that all the space-based targets are met for each feature and each attribute space. For each feature and attribute space, the total weighted distance between the demand points and their closest selected planning units is calculated ($\lambda_{fsi} d_{fsij} Y_{fsij}$). This total weighted distance is then scaled by the inherent variation in the demand points ($\lambda_{fsi} \delta_{fsi}$). The resulting fraction yields a proportion conceptually similar to the statistical $R^2$ value, and the constraints ensure that this proportion must be greater than or equal to the space-based target. Constraints (3d) ensure that only one planning unit is assigned to each demand point. Constraints (3e) ensure that demand points are only assigned to selected planning units. Constraints (3f) ensure that the $X$ and $Y$ variables are binary.

### Reliable formulation
The reliable formulation explicitly considers the probability that the planning units are inhabited. As a consequence, it may deliver prioritisations that will sufficiently represent an attribute space even if the features do not inhabit several of the planning units when the prioritisation is implemented. This behaviour is achieved by siting back-up planning units near selected planning units with low occupancy probabilities in the attribute space(s). To ensure that prioritisations are robust against multiple planning units being uninhabited, the problem assigns planning units at multiple backup levels.

Backup levels levels are defined as $r$-levels [similar to failure levels in @r18]. The first backup $r$-level is used to calculate the level of representation when all of the selected planning units are occupied by all $f \in F$. For this scenario, the closest selected planning unit to each demand point $i$ for attribute space $s$ is assigned at $r$-level$=0$. This scenario essentially represents $Y_{fsij}$ in the unreliable formulation. The second backup $r$-level is used to assess the level of representation when the closest planning unit to each demand point $i$ is unoccupied. For this scenario, the second closest planning units are assigned at $r$-level$=1$. The third backup $r$-level is used to assess representation when the first two closest planning units are unoccupied. The third closest planning units are assigned at $r$-level$=2$. Continuing on, in this manner, the selected planning units in a prioritisation are assigned to each demand point $i \in I$, attribute space $s \in S$, and each feature $f \in F$ at an $r$-level.

A final backup $r$-level when $r=R$ is used to assess the level of representation when the features $f \in F$ do not occupy any selected planning units in a prioritisation. Each demand point $i \in I$ for each $s \in S$ and $f \in F$ is assigned to an "imaginary" planning unit $j=J$ at $r=R$. The distance variables associated with this imaginary planning unit $d_{fsiJ}$ denote the loss of biological value associated with failing to secure a representative sample of feature $f$ in attribute space $s$. However, the $d$ variables are in distance units which are meaningless in this context. Thus these variables are calculated using a failure multiplier ($M$) and the maximum distance between the planning units and the demand points for $f \in F$, $s \in S$ (4).

\begin{align*}
& d_{fsiJ} = M \max\limits_{0 \leq i \leq I-1, 0 \leq j \leq J-1} d_{fsij} & \forall & 0 \leq f \leq F-1, \tag{4} \\
& & & 0 \leq s \leq S-1\\
\end{align*}

Moderately-sized conservation planning problems often include several thousand planning units. It is currently not be feasible to solve this problem when considering all possible failure scenarios. As a consequence, the $R$ variable can be any $1 \leq R \leq J-1$. For instance, when $R=3$ only 2 backup levels are considered in addition to the final backup level. Cui et al. [-@r16] found that $R=5$ yields similar solutions to $R=J$ when $J >> 5$. However, in most costs the decision maker will likely be limited to $R=1$ to obtain prioritisations in a feasible amount of time.

In the reliable formulation, the control variables are the $B$ (1a), $T_{s}$ (1b), $\tau_{sa}$ (1c), $R$, and $M$ variables.

\begin{align*}
R &= \parbox{25em}{number of failure levels} \tag{5a} \\
%
M &= \parbox{25em}{failure multiplier} \tag{5b}\\
%
\end{align*}

The decision variables are the $X_j$ (2a), $Y_{fsijr}$, $P_{fsijr}$ variables.

\begin{align*}
Y_{fsijr} &= \begin{cases}
		1, & \parbox{25em}{if demand point $i$ is assigned to planning unit $j$ for feature $f$ in space $s$ at back-up level $r$. } \tag{6a} \\
		0, & \parbox{25em}{otherwise} \\
	\end{cases} \\
%
P_{fsijr} &= \parbox{25em}{probability that demand point $i$ is assigned to planning unit $j$ at back-up level $r$ for feature $f$ and space $s$} \tag{6a}\\
%
\end{align*}

The reliable formulation (RRAP) is a multi-objective optimisation problem.

\begin{align*}
& \text{(RRAP)} & \text{Min } & \text{(3a)}\\
%
& & \text{s.t. } & \text{(3b)}\\
%
& & & 1 - \frac{\sum_{i=0}^{I-1} \sum_{j=0}^{J-1} \left( \lambda_{fsi} d_{fsij} P_{fsijr} Y_{fsij} \right)^2}{\sum_{i=0}^{I-1} \left( \lambda_{fsi} \delta_{fsi} \right)^2 } \geq T_{fs}  & \forall & 0 \leq f \leq F-1, \tag{7a}\\
& & & & & 0 \leq s \leq S-1\\
%
& & & \sum_{j=0}^{J-1} Y_{fsijr} = 1 & \forall & 0 \leq f \leq F-1, \tag{7b}\\
& & & & & 0 \leq s \leq S-1,\\
& & & & & 0 \leq i \leq I-1,\\
& & & & & 0 \leq r \leq R\\
%
& & & \sum_{r=0}^{R} Y_{fsijr} = 1 & \forall & 0 \leq f \leq F-1, \tag{7c}\\
& & & & & 0 \leq s \leq S-1,\\
& & & & & 0 \leq i \leq I-1,\\
& & & & & 0 \leq j \leq J\\
%
& & & \sum_{r=0}^{R-1} Y_{fsijr} \leq X_j & \forall & 0 \leq f \leq F-1, \tag{7d}\\
& & & & & 0 \leq s \leq S-1,\\
& & & & & 0 \leq i \leq I-1,\\
& & & & & 0 \leq j \leq J-1\\
%
& & & Y_{fsiJR} = 1 & \forall & 0 \leq f \leq F-1, \tag{7e}\\
& & & & & 0 \leq s \leq S-1,\\
& & & & & 0 \leq i \leq I-1\\
%
& & & P_{fsij0} = q_{fj} & \forall & 0 \leq f \leq F-1, \tag{7f}\\
& & & & & 0 \leq s \leq S-1,\\
& & & & & 0 \leq i \leq I-1,\\
& & & & & 0 \leq j \leq J\\
%
& & & P_{fsijr} = \left(1 - \right) \sum_{k=0}^{J-1} \frac{1 - q_k}{q_k} P_{f,s,i,k,r-1} Y_{f,s,i,k,r-1}  & \forall & 0 \leq f \leq F-1, \tag{7g}\\
& & & & & 0 \leq s \leq S-1,\\
& & & & & 0 \leq i \leq I-1,\\
& & & & & 0 \leq j \leq J,\\
& & & & & 1 \leq r \leq R\\
%
& & & X_j, Y_{fsijr} \in {0,1} & \forall & 0 \leq f \leq F-1, \tag{7h}\\
& & & & & 0 \leq s \leq S-1,\\
& & & & & 0 \leq i \leq I-1,\\
& & & & & 0 \leq j \leq J,\\
& & & & & 0 \leq r \leq R\\
\end{align*}

The objective function for the reliable formulation is the same as for the unreliable formation (3a). Similar to the unreliable formulation, constraints (3b) and (7a) ensure that the amount-based and space-based targets are met. Constraint (7b--7c) ensure that each planning unit is only assigned to one backup $r$-level for $i \in I$. Constraints (7d) ensure that only selected planning units are assigned to demand points $i \in I$. Constraints (7e) ensure that the imaginary planning unit is always assigned to the highest backup $r$-level. Constraints (7f--7g) determine the probability that planning unit $j$ will be used to sample demand point $i \in I$ for $s \in S$ and $f \in F$ [see @r16 for more information]. Constraints (7h) ensure that the $X$ and $Y$ variables are binary.

### Optimisation
The unreliable and reliable formulations are non-linear. However, the non-linear components can be linearised using existing techniques. First, the expression $X_j X_k$ in (3a) can be linearised using methods described by Beyer et al. [-@r426]. Second, the expression $P_{fsijr} Y_{jsijr}$ in (7a) can be linearised using techniques described by Sherali and Alameddine [-@r20] as implemented in Cui et al. [-@r16]. Linearised versions of the problems can be solved using commercial exact algorithm solvers.

The \texttt{rapr} R package provides functions to express conservation planning data as an optimisation problems using linearised versions of the unreliable and reliable formulations. These optimisation problems can then be solved to generate prioritisations using the commercial \texttt{Gurobi} software suite (\url{http://www.gurobi.com}). Presently, academics can obtain a license at no cost from the Gurobi website. After installing the \texttt{Gurobi} software suite, users will need to install the \texttt{Gurobi R} package. This \texttt{R} package can be installed on recent Windows, Mac OSX, and Ubuntu Linux operating systems.

To understand the beaviour of this problem and showcase its utility, we conducted a simulation study and two case-studies. These studies involved generating prioritizations using only amount-based targets to represent prioritisations generated using conventional methods (eg. Marxan), and prioritizations using amount-based and space-based targets using the unreliable formulation.

## Simulation study
### Methods
We generated a study area with planning units arranged in $`r sqrt(sim.params.LST[[MODE]]$number.planning.units)` \times `r sqrt(sim.params.LST[[MODE]]$number.planning.units)`$ square grid (Figure 2). We then simulated three species. The first species was simulated to represent a hyper-generalist--occurring in all planning units with equal probability (Figure 2a). This species' distribution was based on a uniform distribution (eqn 8a). The second species was simulated to represent a species with an idealised distribution (Figure 2b). It has a core range area, and is less likely to be found in areas that are distant from the core area. This species' distribution was simulated using the probability density function of a single multivariate normal distribution (eqn 8b). The third species represents a species with two distinct populations (Figure 2c). This species' distribution was simulated to depict a bimodal distribution, based on the combination of the probability density functions of two multivariate normal distributions (eqn 8c). The probability that each species inhabited a given planning unit was calculated using the $X, Y$ coordinates of the planning unit and eqns 8a--8c. We used a geographic attribute space to showcase the behavior of the problem. For each species, demand points were generated by calculating the centroids of each planning units and their weights were set as the probability that the planning units were occupied.

@@

We generated four prioritisations for each species. First to represent prioritisations generated using conventional conservation planning methods, we generated prioritisations using only `r sim.params.LST[[MODE]]$amount.target*100` % amount targets. Second to show how the addition of geographic targets can affect a prioritisation, we  generated prioritisations using `r (sim.params.LST[[MODE]]$amount.target*100)` % amount targets and `r (sim.params.LST[[MODE]]$space.target*100)` % geographic targets. Third to represent prioritisations using conventional planning methods that penalise for fragmentation, we generated prioritisations using `r (sim.params.LST[[MODE]]$amount.target*100)` % amount targets and a boundary length modifier of `r sim.params.LST[[MODE]]$blm`. Fourth to illustrate the combined affect using geographic targets and penalising fragmentation, we generated prioritisations using using `r (sim.params.LST[[MODE]]$amount.target*100)` % amount targets and `r (sim.params.LST[[MODE]]$space.target*100)` % geographic targets and a boundary length modifier of `r sim.params.LST[[MODE]]$blm`.

@@

## Case-study 1
### Methods
Here we investigated how space-based targets can be used in a multi-species planning context to generate a prioritisation that sufficiently preserves the realised niche for several species. We selected Queensland, Australia as the study area. This region is ideal for exploring the potential of niche-based targets because it contains a broad range of different habitats. We obtained data for 19 bioclimatic variables across the region [at $30^{\prime \prime}$ resolution from \url{www.worldclim.org}; @r33] and subject them to a principal components analysis (using ArcMap 10.3.1). We used the first `r as.character(as.english(cs1.params.LST[[MODE]]$n.pc.layers))` principle components (cumulatively explained `r round(cs1.pca.DF[[4]][cs1.params.LST[[MODE]]$n.pc.layers],1)` % of the total variation) to characterise the environmental variation across the study area (Figure 3).

We used four bird species in this case study: blue-winged kookaburra (_Dacelo leachii_), brown-backed honeyeater (_Ramsayornis modestus_), the brown falcon (_Falco berigora_), and the pale headed rosella (_Platycercus adscitus_). We used these bird species because they have different evolutionary histories and life-history strategies. We then mapped the extent of occurrence for each species using the following method: we obtained observation data from the Atlas of Living Australia across the whole of Australia [using the \texttt{ALA4R R} package; @r453], spatially thinned the data to omit points within `r cs1.params.LST[[MODE]]$thin.distance/1000` km$^2$ of each other to ameliorate the effects of sampling bias [using the developmental version of the \texttt{spThin R} package; \url{www.github.com/mlammens/spThin}; @r454], and fit `r cs1.params.LST[[MODE]]$mcp.percent` % minimum convex polygons [using the \texttt{adehabitatHR R} package; @r455]. We used this method because it is reproducible using freely available data and moderately accurate.

We generated `r cs1.params.LST[[MODE]]$dp.number` demand points for each species (Figure S2). To achieve this, for each species, we generated random points inside the species range and at each point extracted the principal component values at that location. We then fit hyperbox kernels to the distribution of these points to characterise the realised niche of each species [using a manually chosen bandwidth of `r cs1.params.LST[[MODE]]$dp.bandwidth[1]` and a `r cs1.params.LST[[MODE]]$dp.quantile` quantile to map the core parts of the species' niches; implemented in the \texttt{hypervolume R} package; @r231]. We then generated uniformly distributed points inside the species' kernels and estimated the density of the training points at the uniformly generated points. These uniformly distributed points and associated density estimated were used as demand point coordinates and weights (respectively).

We generated two prioritisations. The first prioritisation was generated using `r cs1.params.LST[[MODE]]$amount.target*100` % amount-based targets for all species. The second prioritisation was generated the same amount-based target with an additional `r cs1.params.LST[[MODE]]$space.target*100` % niche-based for each species.

>>>>

---
title: "rapr: Representative and Adequate Prioritisations in R"
author: |
  | Jeffrey O. Hanson$^1$, Jonathan R. Rhodes$^2$, Hugh P. Possingham$^1$, Richard A. Fuller$^1$
  | $^1$School of Biological Sciences, The University of Queensland, Brisbane, QLD, Australia
  | $^2$School of Geography, Planning and Environmental Management, The University of Queensland, Brisbane, QLD, Australia
  | Correspondance should be addressed to jeffrey.hanson@uqconnect.edu.au
date: "`r format(Sys.time(), '%d %B %Y')`"
tags: ["conservation", "biodiversity", "MILP", "reserve-selection", "optimisation", "protected area"]
abstract: "

1. A central aim in conservation is to maximise the long-term persistence of biodiversity. To fulfill this aim, reserve networks are used to safeguard biodiversity patterns (eg. species, populations) and processes (eg. evolutionary processes). Reserve selection is often formulated as an optimisation problem to identify cost-effective solutions. These formulations often use \"amount targets\" to specify the minimum amount habitat that must be preserved for each species. Although such targets are well suited for securing biodiversity patterns, they are not well suited for obtaining a representative sample of the biodiversity processes that play out across species' ranges. Currently, decision makers lack the tools to develop prioritisations that preserve both biodiversity patterns and processes.

2. To address this issue, we developed the \\texttt{rapr R} package. This \\texttt{R} package provides a toolkit to guide reserve selection using explicit targets for biodiversity processes. Using a novel formulation of the reserve selection problem, users can set \"space targets\" to secure a representative sample of each species within an attribute space to preserve biodiversity processes (eg. an attribute space based on genetic data to preserve evolutionary processes). We explored the functionality of this \\texttt{R} package using a simulation study and two case studies. In each study, we generated prioritisations using amount targets--to represent conventional reserve selection methods--and compared them to prioritisations generated using both amount and space targets.

3. We demonstrated how explicitly considering biodiversity processes in reserve selection can affect the resulting solution. The results from the simulation study suggest that most species could benefit from the explicit use of space-based targets--especially species with multi-model distributions in an attribute space. The results from both first case studies demonstrated that explicitly considering space-based targets for biodiversity processes can result in more effective prioritisations.

4. The \\texttt{rapr R} package provides a unified framework for biodiversity patterns and species-specific processes in conservation planning. Overall, we found strong evidence that prioritisations developed  using amount targets may fail to secure  that can provide better outcomes than just considering biodiversity patterns.
"
output:
  rmarkdown::pdf_document:
    toc: false
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: preamble-latex.tex
fontsize: 11pt
documentclass: article
bibliography: references.bib
csl: reference-style.csl
vignette: >
  %\VignetteIndexEntry{rapr}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

@@

# Introduction
<!-- conservation and reserve networks -->
The overarching aim of conservation is to maximise the long-term persistence of biodiversity [@r423; @r255]. To achieve this, conservation actions must preserve biodiversity patterns (eg. species, populations) and the processes that sustain them. One of the major tangible achievements of modern conservation has been the act of setting aside areas for preservation [@r440]. Reserve networks buffer species from threatening processes [eg. urbanisation; @r255] and set the stage for direct management interventions [eg. captive breeding and reintroduction programs; @r290]. However, the resources available for conservation action are limited, and so reserve networks must be sited in places that satisfy conservation objectives for minimum cost [@r255]. To achieve this, reserve selection is often formulated as an optimisation problem and then solved to identify cost-effective candidate reserve systems [prioritisations; @r255].

<!-- biodiversity processes are important -->
To fulfill the overarching aims of conservation, reserve networks must preserve both ecological and evolutionary processes [@r3; @r255]. Ecological processes, such as predator-prey interactions, pollination, and decomposition, are required for biodiversity to persist over short time-scales. Typically, they operate over small geographic domains--with exceptions such as migration and refugial habitats--and can be preserved using suitably large planning units [@r446] that each contain a discrete unit of habitat [@r448]. On the other hand, evolutionary processes are required for biodiversity to persist over long time-scales, and they typically operate over large geographic domains. Protected areas must preserve adaptive evolutionary processes to foster resilience against environmental change [eg. climate change; @r247]. Protected areas must also preserve neutral evolutionary processes [@r5; @r67; @r63]. These processes are associated with the breakdown of gene flow between populations. They are important for maintaining genetic diversity, and avoiding inbreeding depression. In recent decades, a wealth of data on biodiversity processes has become available to conservation planners [eg. bioclimatic, genetic, and trait data; @r33; @r453; @r460]. Yet this data is only rarely used to guide conservation planning [@r61]. This is due to that fact that existing reserve selection tools focus on preserving biodiversity patterns or processes--but not both.

@@

<!-- aims of this paper -->
Today, one of the key issues in reserve selection is the lack of a unifying decision support tool that can accommodate data on biodiversity patterns and processes in a multi-species context. To fill this void, we present the \texttt{rapr R} package. This \texttt{R} package uses novel formulations of the reserve selection problem to provide decision makers with the tools to generate prioritisations that preserve biodiversity patterns and processes. We aim to define the concepts behind the problem formulations. Furthmore, we aim to explore the functionality of the \texttt{R} package by applying one of the formulations to a set of simulated species and two case studies.

# Methods
## PROBLEM FORMULATION
Biodiversity features are defined as the entity(s) that the prioritisation is required to preserve (eg. species, populations). Spatial attributes are defined as the variation across the species' range that the prioritisation is required to sample. They can be intrinsic (eg. genetic or trait variation) or extrinsic (eg. environmental variation) to the feature. These attributes should be related to the biodiversity processes that the decision maker aims to preserve.

@@

To illustrate these concepts, we will briefly describe a conservation planning scenario example involving an attribute spaces and demand points. A decision maker may wish to develop a prioritisation for a single species. This species has four populations in the study area. However, the decision maker can only afford to preserve three population. The decision maker needs to preserve the adaptive landscape of the species, and preserve populations inhabiting different environmental conditions. To describe environmental variaiton, the decision maker obtained data on the environmental conditions (rainfall (ml) and temperature ($^{\circ}$C)) where each population was found. The decision maker then used this environmental data to construct a two-dimensional environmental attribute space. Next, the decision maker generated demand points as equi-distant points between the range of values where the populations were found. By comparing the distribution of the demand points to the distribution of the populations in the attribute space, the decision maker can identify a suitable prioritisation (Fig. 1). We can see that if the decision maker preserves both populations $A$ and $C$, they will effectively "double-up" on the same environmental characteristics, and in turn their waste resources. Instead, a more representative sample of the intra-specific variation could be preserved by securing populations $A$, $B$, and $D$. This example demonstrates how the inclusion of biodiversity processes can guide the reserve selection process.

The formulations used to express the reserve selection problem in the \texttt{rapr R} package are based on a combination of the \texttt{Marxan} reserve selection problem and the uncapacitated facility location problems [@r451]. Although the \texttt{rapr R} package provides two novel formulations, for brevity, we will define the simpler formulation--referred to as the unreliable formulation--below and define the more complex version--the reliable formulation--in the Supporting Information. The key difference between these two formulations is that the reliable formulation explicitly considers the probability that planning units are occupied when calculating the proportion of an attribute space sampled in a solution. All mathematical terms defined hereafter are described in Table S1. For convenience, the cardinality of sets will be denoted using the same symbol used to denote the variable.

Define $F$ to be the set of features (indexed by $f$). Let $J$ be a set of planning units (indexed by $j$). Also, let $A_j$ denote the area, and $C_j$ denote the cost of preserving planning unit $j \in J$. To assess the extent to which each feature is secured in a given prioritisation, let $q_{fj}$ denote the probability of feature $f$ occupying planning unit $j$. The level of fragmentation associated with a prioritisation is parametrised as the net exposed boundary length. Let the shared edges between each planning unit $j \in J$ and $k \in J$ be $e_{jk}$.

@@

The control variables for the unreliable formulation are the $B$, $T_{s}$, and $\tau_{sa}$ variables.

@@

The objective function (3a) determines the utility of a given prioritisation: a combination of the total cost of a prioritisation and how fragmented it is. Constraints (3b) ensure that all the amount-based targets are met. Constraints (3c) ensure that all the space-based targets are met for each feature and each attribute space. For each feature and attribute space, the total weighted distance between the demand points and their closest selected planning units is calculated ($\lambda_{fsi} d_{fsij} Y_{fsij}$). This total weighted distance is then scaled by the inherent variation in the demand points ($\lambda_{fsi} \delta_{fsi}$). The resulting fraction yields a proportion conceptually similar to the $R^2$ statistic used in $k$-means clustering analysis. The constraints ensure that this proportion must be equal to or greater than the space-based target. Constraints (3d) ensure that only one planning unit is assigned to each demand point. Constraints (3e) ensure that demand points are only assigned to selected planning units. Constraints (3f) ensure that the $X$ and $Y$ variables are binary.

## OPTIMISATION
The unreliable formulation is non-linear. However, the non-linear components can be linearised using existing techniques. The expression $X_j X_k$ in (3a) can be linearised using methods described by Beyer et al. [-@r426]. Linearised versions of the problems can be solved using commercial exact algorithm solvers. The \texttt{rapr} R package provides functions to express conservation planning data as optimisation problems using linearised versions of the unreliable and reliable formulations. These optimisation problems can then be solved to generate prioritisations using the commercial \texttt{Gurobi} software suite (\url{http://www.gurobi.com}). Presently, academics can obtain a license at no cost from the Gurobi website. After installing the \texttt{Gurobi} software suite, users will need to install the \texttt{Gurobi R} package.

# Examples

To understand the behaviour of the unreliable problem and showcase its value, we conducted a simulation study and two case studies. These studies involved generating solutions using only amount targets to represent prioritisations generated using conventional methods (eg. \texttt{Marxan}), and solutions using amount and space targets using the unreliable formulation. By comparing these solutions, we can guage the benefits of explicitly including space targets in reserve selection. All analyses were performed in \texttt{R} [version `r getRversion()`; @r192].

## SIMULATION STUDY
### Methods
We simulated a hypothetical study area with planning units arranged in $`r sqrt(sim.params.LST[[MODE]]$number.planning.units)` \times `r sqrt(sim.params.LST[[MODE]]$number.planning.units)`$ square grid (Fig. 2). We then simulated three species across this study area. The first species was simulated to represent a hyper-generalist--occurring in all planning units with equal probability (Fig. 2a). This species' distribution was based on a uniform distribution (eqn 8a). The second species was simulated to represent a species with an idealised distribution (Fig. 2b). It has a core range area, and is less likely to be found in areas that are distant from the core area. This species' distribution was simulated using the probability density function of a single multivariate normal distribution (eqn 8b). The third species represents a species with two distinct populations (Fig. 2c). This species' distribution was simulated to depict a bimodal distribution, based on the combination of the probability density functions of two multivariate normal distributions (eqn 8c). The probability that each species inhabited a given planning unit was calculated using the $X, Y$ coordinates of the planning unit and eqns 8a--8c. We used a geographic attribute space to showcase the behavior of the problem. For each species, demand points were generated by calculating the centroids of each planning units and their weights were set as the probability that the planning units were occupied.

@@

We generated four solutions for each species. First to represent solutions generated using conventional conservation planning methods, we generated solutions using only `r sim.params.LST[[MODE]]$amount.target*100` % amount targets. Second to show how the addition of geographic targets can affect a prioritisation, we generated solutions using `r (sim.params.LST[[MODE]]$amount.target*100)` % amount targets and `r (sim.params.LST[[MODE]]$space.target*100)` % geographic targets. Third to represent solutions using conventional planning methods that penalise for fragmentation, we generated solutions using `r (sim.params.LST[[MODE]]$amount.target*100)` % amount targets and a boundary length modifier of `r sim.params.LST[[MODE]]$blm`. Fourth to illustrate the combined affect using geographic targets and penalising fragmentation, we generated solutions using using `r (sim.params.LST[[MODE]]$amount.target*100)` % amount targets and `r (sim.params.LST[[MODE]]$space.target*100)` % geographic targets and a boundary length modifier of `r sim.params.LST[[MODE]]$blm`.

@@

The uniform species was simulated with a constant probability of occupancy across the study area (Fig. 2a). The solution generated using `r sim.params.LST[[MODE]]$amount.target*100` % amount-based targets (Fig. 3a) selected `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$n` planning units near the southern end of study area. This configuration is an artifact of the method used to solve this particular instance of the reserve selection problem. In fact, for this species, all prioritisations containing `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$n` planning units are optimal when considering only `r sim.params.LST[[MODE]]$amount.target*100` % amount targets. In the absence of criteria to guarantee representativeness, reserve selection methods may or may not return solutions that secure a representative sample of the features. In this particular case, the solution does not secure a representative sample of the species' range (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount\ntargets')$space.held,2)` % sampled).

The addition of a `r sim.params.LST[[MODE]]$space.target*100` % geographic target resulted in a solution that secured a representative sample of the uniform species' range (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount &\nspace targets')$space.held,2)` % sampled; Fig. 3b). Since all planning units have cost the same amount and have an equal chance of being occupied, this solution has the same number of planning units as the solution generated using only amount targets (cf. Fig 3a). Although the use of amount and space based targets has addressed adequacy and representativeness objectives (respectively), they have resulted in a solution that is a highly fragmented solution.

The addition of a positive boundary length modifier (BLM) parameter resulted in a well-connected solution that secured an adequate proportion (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount &\nspace targets')$amount.held,2)` % ) and representative sample (`r round(filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount & space\ntargets and BLM')$space.held,2)` % sampled) of the species geographic distribution (Fig. 3d). This solution contains `r filter(sim.spp.DF, species=='Uniform\nspecies', prioritisation=='Amount & space\ntargets and BLM')$n` planning units--a few more than the previously discussed solutions--to ensure that the solution secures a representative proportion of the species geographic distribution in a configuration that is not highly fragmented. This result suggests that the combination of space targets and boundary length modifiers may yield solutions that contain more planning units, since the solution generated using amount targets and a boundary length modifier (Fig. 3c) contained the sample number of planning units as the solution with just the amount targets (Fig. 3a). Overall, these results show that under the simplest of conditions, the reliance on just amount targets can cause an under-specified reserve selection problem that is unlikely to return a suitable solution for implementation.
